---
title: "Week 3 Lab: More than Two Samples"
author: "E"
date: "2/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Eco-636/Week 3")
```

## ANOVA Background
General Linear Model with a continuous response variable, **y**, and a singel categorical explanatory variable with **g** groups or levels
In mathematical form:

$$ y_{i} = \beta_0 + \beta_{1(g)}X_{i(g)} + e_i$$
Lets say we have three groups: Group 1, Group 2, and Group 3, with Group 1 being the reference group. The model would look like: 

$$ y_i = \beta_0 + \beta_{1(2)}X_{2i} +  \beta_{1(3)}X_{3i} + e_i $$

### Lets put it into practice
Generating a synthetic data set:
```{r mydata}
my_dataset <- data.frame("skull" = rnorm(6, 10, 2), 
                         "group" = factor(rep(c('A', 'B', 'C'), each = 2)))
my_dataset
```

What does the model matrix look like? 
```{r}
glm(skull ~ group, data = my_dataset)
```
Lets break down that output. 
  - Intercept: $\beta_0$ - represents the mean of the reference group (Group 1)
