---
title: "Week 5 Lab Report"
author: "E"
date: "3/8/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Eco-636/Week 5")
```

## Interacting with Migrants

> Goal: Recommend habitat purchases of either anthro or natural sites for migrant birds. 

Write a brief summary describing findings and recommendations describing your findings and recommendations (no more than 2 pages of single spaced writing, not including code or figures). Summary should include your process and results, including 1-2 supporting figures for model interpretation/prediction, and 1-2 figures for evaluating model assumptions. 

```{r load data}
migrants <- read.csv(file = "Data/migrants.csv", header = TRUE)
str(migrants)
head(migrants)

# We notice that our data includes both migrants and resident birds. Lets subsect for only the migrants
migrants <- subset(migrants, migrants$migstat == "mig", c('migstat','urb.cat', 'n.obs'))
table(migrants$migstat)
summary(migrants)
```

### STEPS
> Develop a set of hypotheses (null and alternative) based on project goals

1. State the question
    * Is there a significant difference in number of migrant birds among antho and natural sites?
    * Response:
        * Number of birds observed
    * Explanatory:
        * sites - `anthro` or `nat`
2. Data Exploration
```{r data explr}
boxplot(migrants$n.obs, ylab = "Migrants")
boxplot(migrants$n.obs ~ migrants$urb.cat)

tapply(migrants$n.obs, migrants$urb.cat, mean) # Raw means
```
So we see a difference in the raw means but is it significant??? 

> Describe all relevant statisical model(s) in words and algebra

3. Describe the model:
    
    *In words:
        * Is there a difference between the habitat means of migrants observed?
        * $H_0$: There is no difference
    * In Mathematical form
        * $y_i = beta_0 + beta_{1(g)}Site_{1i(g)} + e_i$
        * $y_i$ is number of migrants observed
    * Model assumptions are:
        * Residuals are normally distributed
        * Contant variance (homogeneity)
        * Observations are independent
        * Predictors measured without error (fixed X)

> Fit candidate models and evaluate using AIC to select best candidate model

```{r null model}
m0 <- lm(n.obs ~ urb.cat, data = migrants)
summary(m0)
```

> Evaluate and validate the top model(s)

```{r eval models}
par(mfrow = c(3,1))
plot(m0)
```

> Interpret results, including description of all model parameters and what estimates mean including graphical and verbal summaries of the model predictions

```{r}
summary(m0)
```


> Include an annotated R script or do this as an R Markdowon file
